Working title: Ecological network control

### 2018-05-18 paper/manuscript.Rmd
Better explain matchings

A system is said to be controllable if it is possible to steer it from an initial to an arbitrary final state within finite time [@kalman_mathematical_1963]. 
A simple dynamic system can be described by $\frac{dx}{dt} = Ax + Bu(t)$, where the change of its state over time ($\frac{dx}{dt}$) depends on its current state *x* (for example, the species' abundances), an external time-varying input *u(t)* (the control signal), and two matrices *A* and *B*, which encode information about the network structure and how species respond to external inputs, respectively. 
In classic control theory, determining wether this system is controllable or not can easily be achieved by checking that its controllability matrix $R= [ \begin{matrix}B & AB & A^{2}B & ...& A^{n-1}B\end{matrix} ]$ has full rank. 
In complex systems, however, employing this rank condition, or numerical approximations of it, is unfeasible because it is hard to fully parameterize *A* and *B* (either the weights of the links change over time or because they are difficult to measures). 
Here we use an approach based on the structural controllability theorem [@lin_structural_1974], which assumes that we are confident that the elements of *A* and *B* have either non-zero or zero values (there is an interaction or not), but that we less sure about the precise magitude of the non-zero values. 
Using this structural approach we can find out the controllability of a system under almost all non-zero parameter realisations.
In addition, a graphical interpretation of the structural controllability theorem enables us to translate the control problem into a topological one: a network is controllable if there are no inaccessible nodes—nodes without incoming links—or dilations—*expansions* of the network (Figure \@ref(fig:fig-example-networks)b; supplementary information section \@ref(S-structural-controllability)).

We are often able to estimate *A* in ecological networks, as it represents the interactions between species. 
Part of the control problem resides on estimating a supportable estimation of *B*, which represents the links between external inputs and species. 
Naively, any ecological community (and any system for that matter) could be controlled if we control the state of every species independently. 
Such approach is typically unpractical. 
We are instead interested in finding a minimum driver node set (effectively finding *B*) with which to make the system controllable. 
However, a brute force search of this minimum driver node set is computationally prohibitive for most networks as it involves the evaluation of $2^N$ different controllability matrices. 
We employ a recently developed approach that shows that the control problem of finding the minimum driver node set can be mapped into a purely graph theoretic problem: maximum matching [@liu_controllability_2011; @liu_control_2016].

Maximum matching is a widely studied topic in graph theory and is commonly used in multiple applications, from dating apps and wireless communications to organ transplant allocation and peer to peer file sharing. 
A matching in an unweighted directed graph is defined as a set of links that do not share a common start or end nodes. 
A node is said to be superior if it is at the beginning of a matched link and called matched if it is at the end. 
Finally, the largest possible matching is called maximum matching. 
These definitions, however, offer little insight about how maximum matching is related to structural controllability. 
The key is to note that these definitions imply that a matched node can have only one matched link pointing to it, and similarly a superior node can have at most one link leaving from it. 
The result is that a matching can actually be decomposed into a set of paths that reveal how a control signal can flow across the links in a network to reach every node that compose it (Figure 1cXX). 
And so, as recently shown [@liu_controllability_2011], the minimum driver node set corresponds exactly to the *unmatched* nodes, those that are not controlled by other nodes in the network.

>>>>

A system is said to be controllable if it is possible to steer it from an initial to an arbitrary final state within finite time [@kalman_mathematical_1963]. 
A simple dynamic system can be described by $\frac{dx}{dt} = Ax + Bu(t)$, where the change of its state over time ($\frac{dx}{dt}$) depends on its current state *x* (for example, the species' abundances), an external time-varying input *u(t)* (the control signal), and two matrices *A* and *B*, which encode information about the network structure and how species respond to external inputs, respectively. 
In classic control theory, determining wether this system is controllable can easily be achieved by checking that its controllability matrix $R= [ \begin{matrix}B & AB & A^{2}B & ...& A^{n-1}B\end{matrix} ]$ has full rank. 
In complex systems, however, employing this rank condition, or numerical approximations of it, is unfeasible because it is hard to fully parameterize *A* and *B* (either because the weight of the links change over time or because they are difficult to measure). 
Here we use an approach based on the structural controllability theorem [@lin_structural_1974], which assumes that we are confident that the elements of *A* and *B* have either non-zero or zero values (there is an interaction or not), but that we less sure about the precise magitude of the non-zero values. 
Using this structural approach we can find out the controllability of a system under almost all non-zero parameter realisations.
Structural controllability also has a graphical interpretation that allow us to translate the control problem into a topological one: a network is controllable if there are no inaccessible nodes—nodes without incoming links—or dilations—*expansions* of the network (Figure \@ref(fig:fig-example-networks)b; supplementary information section \@ref(S-structural-controllability)).

We are often able to estimate *A* in ecological networks, as it represents the interactions between species. 
Part of the control problem resides on estimating a supportable estimation of *B*, which represents the links between external inputs and species. 
Naively, any ecological community (and any system for that matter) could be controlled if we control the state of every species independently. 
Such approach is typically unpractical. 
We are instead interested in finding a minimum driver node set (effectively finding *B*) with which to make the system controllable. 
However, a brute force search of this minimum driver node set is computationally prohibitive for most networks as it involves the evaluation of $2^N$ different controllability matrices. 
We employ a recently developed approach that shows that the control problem of finding the minimum driver node set can be mapped into a graph theoretic problem: maximum matching [@liu_controllability_2011; @liu_control_2016].

Maximum matching is a widely studied topic in graph theory and is commonly used in multiple applications, ranging from dating apps and wireless communications to organ transplant allocation and peer to peer file sharing. 
A matching in an unweighted directed graph is defined as a set of links that do not share a common start or end nodes; the largest possible matching is called maximum matching. 
For example, in a network composed by jobs and job applicants, a matching is any pairing between applicants and positions that satisfies the basic constrain of any applicant being assigned to at most one position and visceversa.
A maximum matching would then be such a pairing that maximises the number of applicants with jobs and the number of positions filled. 
The definition of maximum matching, however, offer little insight about how it is related to structural controllability. 
The key is to note that the conditions of structural controllability imply that there needs to be a one to one relationship between *superior* and *subordinate* nodes (Figure \@ref(fig:fig-example-networks)b, bottom left). 
We use the maximum matching algorithm to pair superior to subordinate nodes in a way consistent with these conditions. 
The result is that a matching can actually be decomposed into a set of paths that reveal how a control signal can flow across the links in a network to reach every node that compose it. 
And so, as recently shown [@liu_controllability_2011], the minimum driver node set, those to which an external control signal should be applied to make the system controllable, corresponds exactly to the *unmatched* nodes in the network (Figure \@ref(fig:fig-control-configuration)).
